Detection of eco-evolutionary dynamics in metacommunities using Joint Species

### 2024-11-29 manuscript/manuscript_v13.Rmd
Edited results, fig1 to svg

Linear regression models can estimate the impacts of intra- and interspecific competition, environmental variation, and trait evolution for population size in simple one- and two-species models of population growth. In simple cases (one-species models, linear and quadratic relationships between environment and population size; models *A* and *B*), the regression coefficients can successfully reflect the direction and magnitude of the relationships included in the non-linear simulation model (the impact of $N_t$ and $E_t$ for $N_{t+1}$). However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure; models *C*, *D*, *E*, *F*), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1).

>>>>

Researchers often use linear regression models to study non-linear population and community dynamics. We evaluated the performance of linear models in simple (1-2 species, 1-few sites) cases, to validate that effect size estimates and predicted densities accurately reflected mechanisms before moving onto more complex models in the next sections of our analysis. When only one species was present (models *A* and *B*), the regression coefficients (model *A*: $N_{t+1} \sim \beta_0 + \beta_1 N_t + \epsilon$; *B* $N_{t+1} \sim \beta_0 + \beta_1 N_t + \beta_2 E_t + \beta_3 E_t^2 + \epsilon$) reflected the direction and magnitude of the mechanistic drivers of population size (i.e. $\beta_1$ correctly captured the positive relationship between $N_t$ and $N_{t+1}$ under our simulation conditions, and $\beta_2$ and $\beta_3$ correctly captured the quadratic unimodal relationship between $E_t$ and $N_t+1$; Appendix S1.1, S1.2).

When a second species was added (models *C* and *D*), the regression coefficients performed well for predicting density values at all time points, but the direction and magnitude of the regression coefficients did not reflect the intra- and interspecific competition strengths included in the mechanistic model (Appendix S1.3). We instead used a latent variable modelling approach (implemented in HMSC) to consider these species interactions as part of the temporal random effects included in the model. This approached produced effect size estimates that both predicted the population dynamics well (Figure 1a \@ref(fig:fig1)) and captured the conditional effects of environment for species densities (Figure 1b \@ref(fig:fig1)). The variation partition also successfully captured the relative importance of the environmental predictor and the temporal random effects (which in this simple case were driven by intra- and interspecific species interactions; Figure 1c \@ref(fig:fig1); Appendix S1.3).

This approach produced estimates of main effects of environment (Figure 1a,b)

However, as the processes included in the simulation models become slightly more complex (competition with a second species, spatial structure; models *C*, *D*, *E*, *F*), regression coefficients were useful for prediction but not for recreating the underlying processes driving changes in population size (Appendix S1).
